---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# shinyDemo

<!-- badges: start -->
<!-- badges: end -->

The goal of `shinyDemo` is to demonstrate my Shiny programming skills to potential
clients by building an app that allows users to interact with some data.

>With this exercise, we want to see your work in analytical and programmatic terms. 
So, we'd like to see what software design you chose for your app. Also, we would
like to know how you structure and robustify your code. Finally, we are keen to 
understand how you employ software quality assurance techniques and version 
control software.
> 
> `r tufte::quote_footer('--- A potential client')`

Among the client's suggestions, I picked to the [Tidy Tuesdays](https://www.tidytuesday.com/) 
`Wealth and income over time` [repository](https://github.com/rfordatascience/tidytuesday/tree/master/data/2021/2021-02-09) 
data set  called `income_distribution.csv`.

## Installation

You can install the development version of `shinyDemo` from [GitHub](https://github.com/) with:

``` {r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("RudyPastel/shinyDemo")
```

## How to use shinyDemo

Obviously, you should try the app out.

```{r example, eval = FALSE}
shinyDemo::startDashboard()
```

More importantly however, read, analyse and evaluate the code and its history.

Comments, suggestions and customer inquiries are welcome. Contact [Dr Rudy Pastel](mailto:rudy.pastelgmail.com).
