---
title: "A first pass at the data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{firstPass}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.asp = 0.8, 
  fig.width = 7, 
  out.width = "100%"
)
```

As the data basis of my app, I picked to the [Tidy
Tuesdays](https://www.tidytuesday.com/) `Wealth and income over time`
[repository](https://github.com/rfordatascience/tidytuesday/tree/master/data/2021/2021-02-09)
data set called `income_distribution.csv`. Before building the app, I need to know 
the data set and how best to represent it. As per the suggestion of the Tidy Tuesday
team, I will focus on enabling visual comparisons. For now, I will exclude statistical
analysis.

## Used packages

I will list here the packages used during the exploration. 
I will load them here for convenience but will no do so in the package code.

```{r}
library(magrittr)
library(ggplot2)
library(ggridges)
library(gganimate)
```



## Data dictionary and a first reflexion 

I decide to trust the Tidy Tuesday team and admit their data is clean.
`income_distribution.csv` is a cleaned version of the the raw data provided by
the [Urban Institute](https://apps.urban.org/features/wealth-inequality-charts/)
and the [US Census](https://www.census.gov/data/tables/time-series/demo/income-poverty/historical-income-households.html).
This file documents households by total money income, race, and 
hispanic origin of householder separated by year and income groups in USA.

|variable            |class     |description |
|:-------------------|:---------|:-----------|
|year                |integer   | Year |
|race                |character | Racial Group |
|number              |double    | Number of households |
|income_median       |integer   | Income median |
|income_med_moe      |integer   | Income median margin of error |
|income_mean         |integer   | Income mean |
|income_mean_moe     |integer   | Income mean margin of error |
|income_bracket      |character | Income bracket (9 total brackets between `<$15,000` and `>$200,000` |
|income_distribution |double    | Income distribution as the percentage of each year/racial group - should add up to 100 for a specific year and race. |

This definition suggests to compare quantities across races over the years. A
typical representation would be to display time series colour coded by race. In 
the case of income_distribution, one would have to represent the whole distribution
for each race fo each year, which may be overwhelming. 

Some variable definition considerations:

* Depending on the plot, one may have to consider the year variable a discrete or continuous.
* There may be some overlap between the races, the definition of which is already 
  questionable.
* The margin of error data may be irrelevant or uninformative to some viewers. 
  Maybe they should only be displayed optionally.

## Download then load the file

The file will be kept as a csv file within the package.

```{r}
filePath = system.file(package = 'shinyDemo', 'extdata', 'income_distribution.csv')
if (filePath == ""){#If there is no file, download it
  filePath = file.path(system.file(package = 'shinyDemo'), 'extdata', 'income_distribution.csv')
  utils::download.file(
    url ='https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/income_distribution.csv',
    destfile = filePath
  )
}
incomeDistribution <- readr::read_csv(file = filePath)
```
## A first look at the file

Let us start with he first 100 rows to understand the structure of the file.

```{r}
head(x = incomeDistribution, n = 100)
```

For a give value of year, only the income_bracket and income_distribution columns
change. This structure may be useful for some statistical analysis but is not
directly suitable for visualisation. I will need to prepare the data before I
display it.

Let us continue with a file summary.

```{r}
summary(incomeDistribution)
```

There are two order of magnitudes between the minimal and the maximal value of number. 
I should enable zooming. Besides, some columns have missing data. Why? 
Are the missing data randomly distributed?

```{r}
image(is.na(incomeDistribution))
```

There seems to be some structure in the NA distribution
 
```{r}
table(incomeDistribution$race)
```

The definition of the races is murky to me. Maybe the they are clear to the
US public. If I have time, I should like the presentation to the ad hoc definition.
Can I use this in the choice of colours?

```{r}
table(incomeDistribution$income_bracket)
```
 
The income bracket were read as characters but are ordered categories / factors.
   
Broken down by year and race, there are three measurements i.e. number,
income_median with income_med_moe and income_mean with income_mean_moe, and one
derivative quantity i.e. income_bracket and income_distribution. This suggests
to make four plots with the possibility to select which race is represented.


## number + year ~ race

Let us represent the number of households by race over the years.

```{r}
# Take the columns you need and format them
numberVsYearByRace = incomeDistribution[, c('number', 'year', 'race')]
numberVsYearByRace = numberVsYearByRace[!duplicated(numberVsYearByRace), ]
summary(numberVsYearByRace)
```
What is the actual missing value?

```{r}
numberVsYearByRace[is.na(numberVsYearByRace$number), ]
```
Why is this value missing?

> Prior to being included in the "Asian" category in the 1980s, many Americans 
> of South Asian descent usually classified themselves as Caucasian or other.
>
> — [Wikipedia](https://en.wikipedia.org/wiki/Asian_Americans)

This should be a reminder that the definition fo these categories evolved over time.
Visual comparison of this data alone will be misleading. I should have some
warnings somewhere.

Let us plot now.

```{r}
rawPlot = ggplot(data = numberVsYearByRace, mapping = aes(x = year, y = number, colour = race)) +
  geom_line() + 
  labs(
    title = 'Number of household by racial group over the years in USA',
    x = 'Year',
    y = 'Number of households',
    colour = 'Racial Group'
  )
rawPlot
```

As expected, the range of the number of households is too large for convenient 
Inspection. Let us represent the race line independently.

```{r}
rawPlot + facet_grid(race ~ ., scales = 'free_y')
```

I need to display it conveniently for the users.


## income_median  + year ~ race

Let us represent the median household income by race over the years.

```{r}
# Take the columns you need and format them
medianIncomeVsYearByRace = incomeDistribution[, c('income_median', 'income_med_moe' , 'year', 'race')]
medianIncomeVsYearByRace = medianIncomeVsYearByRace[!duplicated(medianIncomeVsYearByRace), ]
summary(medianIncomeVsYearByRace)
```



No missing value this time

```{r}
rawPlot = ggplot(data = medianIncomeVsYearByRace, mapping = aes(x = year, y = income_median, colour = race)) +
  geom_line() + 
  labs(
    title = 'Income median by racial group over the years in USA',
    x = 'Year',
    y = 'Income median',
    colour = 'Racial Group'
  )
rawPlot
```

This plot is ok. I definitively need to find a set of meaningful colours for
each race group.


Let us include error margins.

```{r}
# Take the columns you need and format them
rawPlot +
  geom_pointrange(aes(ymin = income_median - income_med_moe, 
                      ymax = income_median + income_med_moe))
```

The error margins decrease as the population size increases. This is of statistical
importance They however do not seem to change the story here. This suggests
to me that they could be omitted on the dashboard.


## income_mean  + year ~ race

Let us represent the mean household income by race over the years.

```{r}
# Take the columns you need and format them
meanIncomeVsYearByRace = incomeDistribution[, c('income_mean', 'income_mean_moe' , 'year', 'race', 'number')]
meanIncomeVsYearByRace = meanIncomeVsYearByRace[!duplicated(meanIncomeVsYearByRace), ]
summary(meanIncomeVsYearByRace)
```



Let us have a look at the missing values.

```{r}
meanIncomeVsYearByRace[is.na(meanIncomeVsYearByRace$income_mean) | is.na(meanIncomeVsYearByRace$income_mean_moe),]
```
We discussed the "Asian Alone" earlier. Wikipedia may help with "White Alone, Not Hispanic".


> In the 2000 Census, people were allowed to check more than one race in addition
  to choosing "Latino". There was strong opposition to this from some civil rights
  activists who feared that this would reduce the size of various racial minorities. 
  The government responded by counting those who are white and of one minority race or 
  ethnicity as minorities for the purposes of civil-rights monitoring and enforcement.
> 
> — [Wikipedia](https://en.wikipedia.org/wiki/Non-Hispanic_whites)

Maybe this data was not properly collected before the 2000 Census. However,
why one would have the income mean *without* margin of error, especially
in the middle of the time series eludes me.


```{r}
rawPlot = ggplot(data = meanIncomeVsYearByRace, mapping = aes(x = year, y = income_mean, colour = race)) +
  geom_line() + 
  labs(
    title = 'Income mean by racial group over the years in USA',
    x = 'Year',
    y = 'Income mean',
    colour = 'Racial Group'
  )
rawPlot
```

I should look for a way to display mean and median together, at least for statisticians.


Let us include error margins.

```{r}
# Take the columns you need and format them
rawPlot +
  geom_pointrange(aes(ymin = income_mean - income_mean_moe, 
                      ymax = income_mean + income_mean_moe))
```

The error margins decrease as the population size increases. This is of statistical
importance They however do not seem to change the story here. This suggests
to me that they could be omitted on the dashboard.


## income_distribution + year ~ race

Here comes the challenge: represent the income distribution by race over the years.

```{r}
# Take the columns you need and format them
incomeDistributtionVsYearByRace = incomeDistribution[, c('income_distribution', 'income_bracket' , 'year', 'race')]
summary(incomeDistributtionVsYearByRace)

orderedBrackets = c(
   "Under $15,000",
   "$15,000 to $24,999",
   "$25,000 to $34,999",
   "$35,000 to $49,999",
   "$50,000 to $74,999",
   "$75,000 to $99,999",
   "$100,000 to $149,999",
   "$150,000 to $199,999",
   "$200,000 and over"
)
incomeDistributtionVsYearByRace$income_bracket = factor(
  x = incomeDistributtionVsYearByRace$income_bracket, 
  levels = orderedBrackets, labels = orderedBrackets, ordered = TRUE)
summary(incomeDistributtionVsYearByRace)

```

Ready to plot 


```{r}
# Plot one year
ggplot(data = incomeDistributtionVsYearByRace[incomeDistributtionVsYearByRace$year == 2019,], 
       mapping = aes(x = income_bracket, 
                     y = income_distribution, 
                     fill = race)) +
  geom_col( alpha = 0.5) + 
  facet_grid(year~race) + 
  labs(
    title = 'Income distribution by racial group in 2019 in USA',
    y = 'Group fraction',
    x = 'Income bracket',
    fill = 'Racial Group'
  )+
  scale_x_discrete(guide = guide_axis(angle = 90)) 
```

This is fine to compare income distribution for one year.

```{r}
# Plot all the years
ggplot(data = incomeDistributtionVsYearByRace[,], 
       mapping = aes(x = income_bracket, 
                     y = income_distribution, 
                     fill = race)) +
  geom_col( alpha = 0.5) + 
  facet_grid(year~race) + 
  labs(
    title = 'Income distribution by racial group over the years in USA',
    y = 'Group fraction',
    x = 'Income bracket',
    fill = 'Racial Group'
  )+
  scale_x_discrete(guide = guide_axis(angle = 90)) 
```

When comparing across time, this is no longer readable. Maybe an animation.

```{r}
# Animate all the years
ggplot(data = incomeDistributtionVsYearByRace, 
       mapping = aes(x = income_bracket, 
                     y = income_distribution, 
                     fill = race)) +
  facet_grid(.~race) + 
  geom_col( alpha = 0.5) + 
  labs(
    title = 'Income distribution by racial group in USA in {floor(frame_time)}',
    y = 'Group fraction',
    x = 'Income bracket',
    fill = 'Racial Group'
  )+
  scale_x_discrete(guide = guide_axis(angle = 90))  + 
  transition_time(year)
  
```
Let us try to stach the distributions vertically.


```{r}
# One year
rawPlot = ggplot(data = incomeDistributtionVsYearByRace[incomeDistributtionVsYearByRace$year == 2019,], 
       mapping = aes(x = unclass(income_bracket), 
                     y = race, 
                     fill = race, 
                     height = income_distribution)) +
  geom_density_ridges2(stat  = "identity", alpha = 0.5) + 
  facet_wrap(year~.) + 
  labs(
    title = 'Income distribution by racial group in 2019 in USA',
    y = 'Group fraction',
    x = 'Income bracket',
    fill = 'Racial Group'
  ) +
  scale_x_continuous(
    labels = levels(incomeDistributtionVsYearByRace$income_bracket),
    breaks = sort(unique(unclass(incomeDistributtionVsYearByRace$income_bracket))), 
    guide = guide_axis(angle = 90)) 
rawPlot

# All the years
rawPlot = ggplot(data = incomeDistributtionVsYearByRace, 
       mapping = aes(x = unclass(income_bracket), 
                     y = race, 
                     fill = race, 
                     height = income_distribution)) +
  geom_density_ridges2(stat  = "identity", alpha = 0.5) + 
  facet_wrap(year~.) + 
  labs(
    title = 'Income distribution by racial group in 2019 in USA',
    y = 'Group fraction',
    x = 'Income bracket',
    fill = 'Racial Group'
  ) +
  scale_x_continuous(
    labels = levels(incomeDistributtionVsYearByRace$income_bracket),
    breaks = sort(unique(unclass(incomeDistributtionVsYearByRace$income_bracket))), 
    guide = guide_axis(angle = 90)) 
rawPlot

```


The second plot is a bit too rich to be showed all at once. I guess an animation 
would make it easier to read.

```{r}
ggplot(data = incomeDistributtionVsYearByRace, 
       mapping = aes(x = unclass(income_bracket), 
                     y = race, 
                     fill = race, 
                     height = income_distribution)) +
  geom_density_ridges2(stat  = "identity", alpha = 0.5) + 
  labs(
    title = 'Income distribution by racial group in the USA in {floor(frame_time)}',
    y = 'Group fraction',
    x = 'Income bracket',
    fill = 'Racial Group'
  ) +
  scale_x_continuous(
    labels = levels(incomeDistributtionVsYearByRace$income_bracket),
    breaks = sort(unique(unclass(incomeDistributtionVsYearByRace$income_bracket))), 
    guide = guide_axis(angle = 90)) + 
  transition_time(year)
```
The user should be able to pick across the global view, the animation or a yearly
view. I prefer to stick to the discrete versions.
