---
title: "A first pass at the data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{firstPass}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

As the data basis of my app, I picked to the [Tidy
Tuesdays](https://www.tidytuesday.com/) `Wealth and income over time`
[repository](https://github.com/rfordatascience/tidytuesday/tree/master/data/2021/2021-02-09)
data set called `income_distribution.csv`. Before building the app, I need to know 
the data set and how best to represent it. As per the suggestion of the Tidy Tuesday
team, I will focus on enabling visual comparisons. For now, I will exclude statistical
analysis.


## Data dictionary and a first reflexion 

I decide to trust the Tidy Tuesday team and admit their data is clean.
`income_distribution.csv` is a cleaned version of the the raw data provided by
the [Urban Institute](https://apps.urban.org/features/wealth-inequality-charts/)
and the [US Census](https://www.census.gov/data/tables/time-series/demo/income-poverty/historical-income-households.html).
This file documents households by total money income, race, and 
hispanic origin of householder separated by year and income groups in USA.

|variable            |class     |description |
|:-------------------|:---------|:-----------|
|year                |integer   | Year |
|race                |character | Racial Group |
|number              |double    | Number of households |
|income_median       |integer   | Income median |
|income_med_moe      |integer   | Income median margin of error |
|income_mean         |integer   | Income mean |
|income_mean_moe     |integer   | Income mean margin of error |
|income_bracket      |character | Income bracket (9 total brackets between `<$15,000` and `>$200,000` |
|income_distribution |double    | Income distribution as the percentage of each year/racial group - should add up to 100 for a specific year and race. |

This definition suggests to compare quantities across races over the years. A
typical representation would be to display time series colour coded by race. In 
the case of income_distribution, one would have to represent the whole distribution
for each race fo each year, which may be overwhelming. Depending on the plot,
one may have to consider the year variable a discrete or continuous.


## Download then load the file

The file will be kept as a csv file wihin the package.

```{r}
filePath = system.file(package = 'shinyDemo', 'extdata', 'income_distribution.csv')
if (filePath == ""){#If there is no file, download it
  filePath = file.path(system.file(package = 'shinyDemo'), 'extdata', 'income_distribution.csv')
  utils::download.file(
    url ='https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/income_distribution.csv',
    destfile = filePath
  )
}
incomeDistribution <- readr::read_csv(file = filePath)
```
## A first look at the file

Let us start with a file summary

```{r}
summary(incomeDistribution)
```
 
A few ideas to keep in mind:
 
 * Some columns have missing data. Why?
 * The margin of error data may be irrelevant or uninformative to some viewers. 
   Maybe they should only be displayed optionally.
   
Broken down by year and race, there are three measurements i.e. number,
income_median with income_med_moe and income_mean with income_mean_moe, and 
derivative quantities i.e. income_bracket and income_distribution. This suggests
to make five plots with the possibility to select which race is represented.


